---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fireexposuR

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)

<!-- badges: end -->

The goal of fireexposuR is to provide a standardized and accessible
platform for the computation and analysis of wildfire exposure. Wildfire
exposure assessments are a decision support tool in wildfire management
and can be applied for multiple temporal horizons and spatial extents.
This package automates the methods previously documented in a series of
scientific publications.

-   [Beverly et al. 2010](https://doi.org/10.1071/WF09071)
-   [Beverly et al. 2021](https://doi.org/10.1007/s10980-020-01173-8)
-   [Beverly and Forbes 2023](https://doi.org/10.1007/s11069-023-05885-3)

The functions in this package require the pre-preparation of data; an
accompanying paper details suggestions for data acquisition and
preparation in accordance with various budget limitations and user
experience levels. The initial release of the package provides a
collection of functions that assist users with conducting wildfire
exposure and directional vulnerability assessments for values and
landscapes, and includes methods for validating the metric for an area
of interest. Outputs from the functions include spatial data, tables,
graphics, and maps that can be further analyzed or modified directly in
R or exported for use in other applications.

## Installation

You can install the development version of fireexposuR from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("heyairf/fireexposuR")
```

## Usage example

This is a basic example which shows a workflow to assess wildfire
exposure on a landscape and within an area of interest.

### Data preperation

First, some example data will be generated:

```{r data}
library(terra)
# generate example hazard data -----------------------------
set.seed(0)
e <- c(45,55,495,505) * 10000
r <- rast(resolution = 100, extent = ext(e))
values(r) <- sample(c(0,1), ncell(r), replace = TRUE)
crs(r) <- "EPSG:32608"
r <- sieve(r, threshold = 50, directions = 4)
haz <- sieve(r, threshold = 500, directions = 4)
# generate example AOI polygon -----------------------------
filepath <- "extdata/builtsimpleexamplegeom.csv"
g <- read.csv(system.file(filepath, package = "fireexposuR"))
m <- as.matrix(g)
aoi <- vect(m, "polygons", crs = haz)
# ----------------------------------------------------------
```

#### Hazard data

The `haz` layer is a binary raster that represents wildland fuels that
are able to generate long-range embers up to a transmission distance of
500 meters with a value of 1.

#### Area of interest

The `aoi` layer is a polygon representing a localized area of interest
(e.g., the built environment of a community, a sensitive habitat, a
campground, etc.) shown in red.

```{r aoivis, echo = FALSE}
plot(haz)
plot(aoi, add = T, col = "red")
```

### Compute exposure

Now, we will use the hazard data to compute the exposure to long-range
ember transmission.

```{r exposureexample}
library(fireexposuR)
exp <- exposure(haz, tdist = "l")
```

### Visualize exposure

Once we have an exposure raster the rest of the package functions can be
used to visualize it in different ways. For a landscape, we can map
exposure with a continuous scale with `mapexpcont()`:

```{r maplandscape}
mapexpcont(exp)
```

We can also see how that exposure is distributed within a localized area
of interest with exposure classifications using `mapexpclass()`. Now we
can see that within our area of interest the northwest corner is a
potential entry point for long-range embers from the landscape.

Note: our imaginary area of interest is in the middle of the Pacific
Ocean so the base map does not provide further reference.

```{r maplocal}
mapexpclass(exp, classify = "local", aoi)
```
