---
title: "Local scale example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{communityscale}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Usage example

This is a basic example which shows a workflow to assess wildfire
exposure to a community using multiple functions from the fireexposuR
package. The package automates the methods within various research
products which can be accessed through the following links:

-   [Beverly et al. 2010](https://doi.org/10.1071/WF09071)
-   [Beverly et al. 2021](https://doi.org/10.1007/s10980-020-01173-8)
-   [Beverly and Forbes
    2023](https://doi.org/10.1007/s11069-023-05885-3)

```{r setup}
library(fireexposuR)
```

### Data preperation

First, some example data will be generated for a town that doesn't exist in the
middle of Alberta.

```{r data}
# load terra for spatial data funtions
library(terra)

# read example hazard data
hazard_file_path <- "extdata/hazard.tif"
hazard <- terra::rast(system.file(hazard_file_path, package = "fireexposuR"))

# read example polygon geometry for area of interest boundary
geom_file_path <- "extdata/builtsimpleexamplegeom.csv"
geom <- read.csv(system.file(geom_file_path, package = "fireexposuR"))

# use geometry to make an area of interest polygon
aoi <- terra::vect(as.matrix(geom), "polygons", crs = hazard)

# generate random points within the aoi to represent values
points <- terra::spatSample(aoi, 200)
```


#### Hazard data

The `hazard` layer is a binary raster that represents wildland fuels that
are able to generate long-range embers up to a transmission distance of
500 meters:

```{r hazardvis, echo = FALSE}
# plot hazard layer with terra defaults
plot(hazard)
```

#### Area of interest and values data

The `aoi` layer is a polygon representing the boundary of a built environment of
this imaginary town. The `points` feature represents values within the
community. We will pretend the points are the locations of structures.

```{r aoivis, echo = FALSE}
# plot the vector data with terra defaults
plot(aoi)
plot(points, add = T)
```

### Compute exposure

Now, we will use the hazard data to compute the exposure to long-range
ember transmission.

```{r computeexposure}
library(fireexposuR)
exposure <- fire_exp(hazard, tdist = "l")
```

### Visualize exposure

Once we have an exposure raster the rest of the package functions can be
used to visualize it in different ways. For a landscape, we can map
exposure with a continuous scale with `fire_exp_map_cont()`:

```{r maplandscape}
fire_exp_map_cont(exposure)
```

We can also see how that exposure is distributed within the built
environment with exposure classifications in an area of interest with
`fire_exp_map_class()`.


```{r maplocal}
fire_exp_map_class(exposure, classify = "local", aoi)
```

This map gives us a better understanding of areas of the town that could
be fire entry points. We can also summarize the area with `fire_exp_summary()` 
if we want to know the proportional distributions of each class.

```{r summlocal}
fire_exp_summary(exposure, classify = "local", aoi)
```

We also have data for the values within the built environment, for which
we can map or summarize in a table as well.

```{r mapvalues}
# extract exposure to the points feature
points_ext <- fire_exp_extract(exposure, points)

# visualize the extracted data in a summary table 
fire_exp_extract_vis(points_ext, classify = "local")

# visualize the extracted data in a map
fire_exp_extract_vis(points_ext, classify = "local", map = TRUE)

```



#### Directional vulnerability

Our make believe town may also wish to assess the directional
vulnerability to wildfire towards their community. This assessment
identifies linear pathways of exposure from the landscape toward a
value.


```{r mapdir, message=FALSE}
# assess directional vulnerability
transects <- fire_exp_dir(exposure, aoi)

# visualize directional vulnerability with a map
fire_exp_dir_map(transects, aoi)

```


