---
title: "communityscale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{communityscale}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Usage example

This is a basic example which shows a workflow to assess wildfire
exposure to a community using multiple functions from the fireexposuR
package. The package automates the methods within various research
products which can be accessed through the following links:

-   [Beverly et al. 2010](https://doi.org/10.1071/WF09071)
-   [Beverly et al. 2021](https://doi.org/10.1007/s10980-020-01173-8)
-   [Beverly and Forbes
    2023](https://doi.org/10.1007/s11069-023-05885-3)

```{r setup}
library(fireexposuR)
```

### Data preperation

First, some example data will be generated for an imaginary town:

```{r data}
library(terra)
# read example hazard data ----------------------------------
filepath <- "extdata/hazard.tif"
haz <- terra::rast(system.file(filepath, package = "fireexposuR"))
# read example AOI
filepath <- "extdata/builtsimpleexamplegeom.csv"
g <- read.csv(system.file(filepath, package = "fireexposuR"))
aoi <- terra::vect(as.matrix(g), "polygons", crs = haz)
# generate random points
pts <- terra::spatSample(aoi, 200)
#' # ----------------------------------------------------------
```

This example data represents the sort of data that would be pre-prepared
by a user for the assessment.

#### Hazard data

The `haz` layer is a binary raster that represents wildland fuels that
are able to generate long-range embers up to a transmission distance of
500 meters:

```{r hazardvis, echo = FALSE}
plot(haz)
```

#### Area of interest and values data

The `aoi` layer is a polygon representing the built environment of our
imaginary town. The `pts` feature represents the centroids of homes and
structures within the community.

```{r aoivis, echo = FALSE}
plot(aoi)
plot(pts, add = T)
```

### Compute exposure

Now, we will use the hazard data to compute the exposure to long-range
ember transmission.

```{r computeexposure}
library(fireexposuR)
exp <- fire_exp(haz, tdist = "l")
```

### Visualize exposure

Once we have an exposure raster the rest of the package functions can be
used to visualize it in different ways. For a landscape, we can map
exposure with a continuous scale with `fire_exp_map_cont()`:

```{r maplandscape}
fire_exp_map_cont(exp)
```

We can also see how that exposure is distributed within the built
environment with exposure classifications in an area of interest with
`fire_exp_map_class()`.

*Note: our imaginary town is in the middle of the Pacific Ocean so the
base map does not provide further reference.*

```{r maplocal}
fire_exp_map_class(exp, classify = "local", aoi)
```

This map gives us a better understanding of areas of the town that could
be fire entry points. We can also summarize the area with `fire_exp_summary()` 
if we want to know the proportional distributions of each class.

```{r summlocal}
fire_exp_summary(exp, classify = "local", aoi)
```

We also have data for the values within the built environment, for which
we can map or summarize in a table as well.

```{r mapvalues}
# extract exposure to the values feature
vals_ext <- fire_exp_extract(exp, pts)
# summary table 
fire_exp_extract_vis(vals_ext, classify = "local")

# map
fire_exp_extract_vis(vals_ext, classify = "local", map = TRUE)

```

With this information, the community has now identified 12 structures
that are extremely exposed to long-range embers from the landscape in
the northwest of the community. This could be a potential area to
prioritize wildfire mitigation strategies.

#### Directional vulnerability

Our make believe town may also wish to assess the directional
vulnerability to wildfire towards their community. This assessment
identifies linear pathways of exposure from the landscape toward a
value.

*Note: our imaginary town is in the middle of the Pacific Ocean so the
base map does not provide further reference.*

```{r mapdir}
transects <- fire_exp_dir(exp, aoi)
fire_exp_dir_plot(transects, aoi, map = TRUE)

```

Now we can see that although the northwest corner of the town is a
potential entry point, the pathway to that location is only viable from
5 kilometers out. The southeast pathway might be a more concerning
pathway because it covers the full 15 kilometers. Depending on local
knowledge, this assessment could identify further areas of concern. For
example, if the region has consistent patterns of southwest winds it may
be a priority area for fuel reduction treatments. Or perhaps there is a
popular outdoor recreation area to the northwest close to the community,
in which case the shorter pathway might be more of a concern if there is
increased human ignition potential in that area.
